---
title: "Getting Started with AAPA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with AAPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**AAPA** (Anchor-Assisted Pedigree Assignment) is an R package for 
high-throughput full-sib family assignment. It is designed for scenarios where:

- Candidate parents are known (enumerable family set)
- Each family has one or more **anchor individuals** with high-confidence 
  family labels
- A large number of test individuals need to be assigned to families (or 
  rejected as unknown)

## Quick Start

### 1. Simulate example data

```{r simulate}
library(aapa)

sim <- simulate_aapa_data(
  n_families = 5,
  n_snps = 200,
  n_offspring_per_family = 10,
  n_anchors_per_family = 2,
  n_unknown = 5,
  missing_rate = 0.01,
  error_rate = 0.001
)

str(sim, max.level = 1)
```

### 2. Prepare input objects

```{r prepare}
geno <- sim$genotype
parents_df <- sim$parents
anchors_df <- sim$anchors

# Build parents object
parents <- lapply(seq_len(nrow(parents_df)), function(i) {
  list(
    family_id = parents_df$family_id[i],
    sire_id   = parents_df$sire_id[i],
    dam_id    = parents_df$dam_id[i],
    sire_geno = geno[parents_df$sire_id[i], ],
    dam_geno  = geno[parents_df$dam_id[i], ]
  )
})
names(parents) <- parents_df$family_id
class(parents) <- "aapa_parents"

# Build anchors object
anchors <- structure(
  anchors_df,
  class = c("aapa_anchors", "data.frame"),
  geno = geno[anchors_df$individual_id, , drop = FALSE]
)
```

### 3. Run quality control (optional)

```{r qc}
qc_result <- qc_filter(geno, max_snp_missing = 0.1,
                       max_ind_missing = 0.2, min_maf = 0.01)
geno_clean <- qc_result$genotype

# If QC removes markers, AAPA will align parents and anchors to the
# filtered marker set by marker name during assignment.
```

### 4. Run family assignment

```{r assign}
result <- aapa_assign(
  genotype = geno_clean,
  parents  = parents,
  anchors  = anchors,
  alpha    = 1.0,
  beta     = 1.0,
  top_k    = 3,
  tau_conf = -0.05,
  tau_rej  = 0.1,
  max_conflict = 0.1
)

print(result)
```

### 5. Evaluate results

```{r evaluate}
# Compare to true labels
asgn <- result$assignment
assigned <- asgn[asgn$status == "ASSIGNED", ]
true_labels <- sim$true_labels

correct <- sum(
  assigned$assigned_family == true_labels[assigned$individual_id],
  na.rm = TRUE
)
cat(sprintf("Accuracy: %.1f%% (%d/%d)\n",
            100 * correct / nrow(assigned), correct, nrow(assigned)))
```

### 6. Visualize results

```{r plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Score distribution
plot_score_distribution(result)

# Rejection diagnostics
plot_rejection_diagnostics(result)
```

## Method Overview

AAPA uses a three-component scoring system:

1. **Mendelian Compatibility**: For each SNP, check whether the test 
   individual's genotype is compatible with the expected offspring 
   genotypes from the candidate parents.

2. **Anchor Kinship**: Compute IBS (Identity-By-State) similarity 
   between the test individual and the family's anchor individuals.

3. **Composite Score**: 
   $$S_{i,f} = -\alpha \cdot C_{i,f} + \beta \cdot K_{i,f}$$
   where $C$ is the conflict rate and $K$ is the kinship score.

### Rejection Rules

Individuals are rejected (labeled `REJECT`) if any of these conditions 
are met:

- Best score below absolute threshold (`tau_conf`)
- Gap between top-1 and top-2 scores too small (`tau_rej`)
- Mendelian conflict rate too high (`max_conflict`)
- All candidate families show high conflict (suspected unknown family)
